Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation 

### 2025-04-08 manuscript/manuscript.qmd
upated manuscript

---
title: "Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation "
author:  
  - name: Erdem M. Karaköylü
    orcid: 0000-0002-6156-1720
    corresponding: true
    email: erdemk@protonmail.com
    roles:
      - Investigation
      - Software
      - Visualization
    affiliations:
      - Consultant
  - name: Susanne E. Craig
    orcid: 0000-0002-8963-0951
    corresponding: true
    email: susanne.e.craig@nasa.gov
    roles:
      - Investigation
      - Project administration
    affiliations:
      - NASA
abstract: Placeholder
date: last-modified
bibliography: references.bib
---

@@

#### Limitations of Existing Approaches
Unfortunately, the development process for traditional ocean color algorithms is founded on a statistical fallacy. Frequentism—the classical approach—assumes that observations are random samples drawn from the studied phenomena. These samples' probabilities are  considered the observed frequencies that capture the the true - albeit noised up and potentially biased - behavior of the system. Under this framework, if a phenomenon is hypothesized to be represented by a model $M$ and data $$D$ is collected, then the model’s representativeness is evaluated solely through the likelihood $p(D∣M)$, which measures the probability of observing the data given that $$M$ is true. The fallacy arises from the mistaken belief that, with sufficiently large amounts of data, one can simply "flip" the likelihood to obtain $p(M∣D)$, thereby yielding a probabilistic statement about $M$. Compounding this error is the imposition of numerous unverified assumptions—disguised as objectivity—that lead frequentists to set the parameters of $M$ by maximizing $p(D∣M)$, under the false premise that this provides the best possible representation of the model. In essence, this fallacy leads to a confusion between the sampling probability $p(D∣M)$ and the inferential probability $p(M∣D)$; the latter  when correctly obtained  quantifies the plausibility of the hypothesis represented by $M$ - precisely what we seek. The historical dimensions of this frequentist illusion are fascinating in their own right, and we invite interested readers to consult works such as McGrayne [@mcgrayne] and Clayton [@clayton].

@@

>>>>

---
title: "Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation "
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:  
  - name: Erdem M. Karaköylü
    orcid: 0000-0002-6156-1720
    corresponding: true
    email: erdemk@protonmail.com
    roles:
      - Investigation
      - Software
      - Visualization
    affiliations:
      - Consultant
  - name: Susanne E. Craig
    orcid: 0000-0002-8963-0951
    corresponding: false
    email: susanne.e.craig@nasa.gov
    roles:
      - Investigation
      - Project administration
    affiliations:
      - NASA
abstract: Placeholder
plain-language-summary: Placeholder
keywords: []
key-points:
  - Key Points convey the main points and conclusions of the article. 
  - Up to three key point statements are allowed, and each is limited to at most 140 characters with no abbreviations.
  - Key Points are not included in the word count.
bibliography: references.bib
citation:
  container-title: Geophysical Research Letters
keep-tex: true
date: last-modified
---

@@

#### Limitations of Existing Approaches
Regrettably, the development process for traditional ocean color algorithms is founded on a statistical erroneous argument. The fallacy is the error of conflating empirical frequency — derived from a limited, noisy set of observations — with the long-run probability that would be obtained under an infinite number of trials,Frequentism—the classical approach—assumes that observations are random samples drawn from the studied phenomena. These samples' probabilities are considered the observed frequencies that capture the true - albeit noised up and potentially biased - behavior of the system. Under this framework, if a phenomenon is hypothesized to be represented by a model $M$ and data $D$ is collected, then the model’s representativeness is evaluated solely through the likelihood $p(D∣M)$, which measures the probability of observing the data given that $M$ is true. The fallacy arises from the mistaken belief that, with sufficiently large amounts of data, one can simply "flip" the likelihood to obtain $p(M∣D)$, thereby yielding a probabilistic statement about $M$. Compounding this error is the imposition of numerous unverified assumptions—disguised as objectivity—that lead frequentists to set the parameters of $M$ by maximizing $p(D∣M)$, under the false premise that this provides the best possible representation of the model. In essence, this fallacy leads to a confusion between the sampling probability $p(D∣M)$ and the inferential probability $p(M∣D)$; the latter  when correctly obtained  quantifies the plausibility of the hypothesis represented by $M$ - precisely what we seek. The historical dimensions of this frequentist illusion are fascinating in their own right, and we invite interested readers to consult works such as McGrayne [@mcgrayne] and Clayton [@clayton].

Regrettably, the development of traditional ocean color algorithms relies on a fundamental statistical error. The error is the mistaken assumption that the relative frequencies observed in a finite sample can be taken as precise estimates of the underlying true probabilities. In other words, it is the error of conflating the empirical frequency—derived from a limited, noisy set of observations—with the long-run probability that would be obtained under an infinite number of trials, thereby ignoring the intrinsic uncertainty and variability that come with finite sampling. The fallout has and continues to be extremely detrimental. Traditional, frequentist approaches, where the variability in sample frequencies is erroneously treated as negligible, yield overconfident inferences, and allegedgly significant findings that do not hold water, in time leading to replication crises.

In the context of statistical modelling, where $M$ is a model that represents a hypothesized relationship between independent and dependent variables, data $D$ is collected to evaluate how well this relationship holds out. A skilled enough model $M$ is retained and used to make predictions; it is put into production. However practical problems arise because under the requentist framework, model evaluation centers on the likelihood $p(D∣M)$, the probability of observing the data given the model is true. With enough data the frequentist assumption is that this likelihood can be directly flipped to give the probability of the model being true, $p(M∣D)$. The leap of faith is completed by imposing unverified assumptions to allow setting model parameters by maximizing $p(D∣M)$. By this logical doing so provides the best model possible. The historical dimensions of how this snake oil was progressively brewed are fascinating in their own right, and we invite interested readers to consult works such as McGrayne [@mcgrayne] and Clayton [@clayton].

@@

## Acknowledgments

## Open research

## References {.unnumbered}

:::{#refs}

:::
